---
title: "example_survival"
format: html
editor: source
---


```{r}
# 1. Install 'pacman' only if it is not already installed on your system
if (!require("pacman")) install.packages("pacman")

# 2. Use pacman to install (if missing) and load all required packages
pacman::p_load(
  survival,    # Core survival analysis algorithms and datasets
  survminer,   # Elegant, publication-ready ggplot2-based survival plots
  gtsummary,   # Publication-ready summary and regression tables
  tidycmprsk,  # Modern, tidy framework for competing risks (Fine-Gray)
  mstate,      # Advanced multi-state modelling and data preparation
  coxme,        # Mixed-effects Cox models for shared frailty/clustering
  ggsurvfit    # More, publication-ready ggplot2-based survival plots
)

# 3. Verify packages are loaded
print("All packages successfully loaded!")
```

inst:	Institution code
time:	Survival time in days
status:	censoring status 1=censored, 2=dead
age:	Age in years
sex:	Male=1 Female=2
ph.ecog:	ECOG performance score as rated by the physician. 0=asymptomatic, 1= symptomatic but completely ambulatory, 2= in bed <50% of the day, 3= in bed > 50% of the day but not bedbound, 4 = bedbound
ph.karno:	Karnofsky performance score (bad=0-good=100) rated by physician
pat.karno:	Karnofsky performance score as rated by patient
meal.cal:	Calories consumed at meals
wt.loss:	Weight loss in last six months (pounds)


```{r}
# Load the built-in lung cancer dataset
data(lung)
lung$sex <- factor(lung$sex, levels = c(1, 2), labels = c("Male", "Female"))

lung
```

```{r}
# 1. Standard Kaplan-Meier
km_fit <- survfit(Surv(time, status) ~ sex, data = lung)

km_fit |> tbl_survfit( probs = c(0.1, 0.25, 0.5, 0.75, 0.9))

```


```{r}

# Publication-ready KM Plot
ggsurvplot(
  km_fit, 
  data = lung, 
  risk.table = TRUE,       # Adds a risk table below the plot
  pval = TRUE,             # Adds log-rank p-value
  conf.int = TRUE,         # Adds confidence intervals
  palette = c("#E7B800", "#2E9FDF"), # Custom colors
  ggtheme = theme_minimal(), # Clean background
  title = "Overall Survival by Sex (Lung Cancer)"
)

 

```

```{r}
# ---------------------------------------------------------
# FIT THE MULTIVARIABLE COX MODEL
# ---------------------------------------------------------
cox_fit <- coxph(Surv(time, status) ~ age + sex, data = lung)

# Publication-ready Regression Table
tbl_regression(cox_fit, exponentiate = TRUE) %>%
  add_global_p() %>%
  bold_p(t = 0.05) %>%
  bold_labels() %>%
  modify_caption("**Table 1. Multivariable Cox Regression Analysis**")
```



```{r}
# 1. Test the PH assumption using Schoenfeld residuals
ph_test <- cox.zph(cox_fit)

# Publication-ready plot of Schoenfeld residuals
ggcoxzph(ph_test, ggtheme = theme_bw(), font.main = 12)
```


```{r}


# ---------------------------------------------------------
# 1. PREPARE DATA FOR SEPARATED PLOTTING
# ---------------------------------------------------------
# Create a "dummy" dataset to separate by sex while holding age constant at its average.
# (Using na.rm = TRUE just in case there are missing age values)
dummy_data <- data.frame(
  sex = c("Male", "Female"), 
  age = c(mean(lung$age, na.rm = TRUE), mean(lung$age, na.rm = TRUE))
)

# Create the survfit object using the dummy data
fit_separated <- survfit(cox_fit, newdata = dummy_data)

# ---------------------------------------------------------
# 2. PUBLICATION-READY ADJUSTED COX PLOT
# ---------------------------------------------------------
ggsurvplot(
  fit_separated, 
  data = dummy_data,                 # Must match the newdata from survfit
  conf.int = TRUE,                   # Adds confidence intervals
  palette = c("#E7B800", "#2E9FDF"), # Custom colors matching your KM plot
  ggtheme = theme_minimal(),         # Clean background
  title = "Adjusted Survival by Sex (Adjusted for Average Age)",
  legend.title = "Sex",
  legend.labs = c("Male", "Female")
)
```


```{r}
# 1. Create a proper 0/1 dummy variable (0 = Male, 1 = Female)
rm(lung); data(lung) # reload the dataset to it s original state
lung$sex_dummy <- lung$sex - 1
lung$sex <- factor(lung$sex, levels = c(1, 2), labels = c("Male", "Female"))

lung
```


```{r}



# 2. Handling Non-PH: Stratified Cox Model
cox_strat <- coxph(Surv(time, status) ~ age + sex, data = lung)

# 3. Handling Non-PH: Time-varying coefficient
cox_time_var <- coxph(Surv(time, status) ~ age + sex + tt(sex_dummy), 
                      data = lung,
                      tt = function(x, t, ...) x * log(t))

# 4. Compare the original model with the time-varying model
tbl_merge(
  tbls = list(
    tbl_regression(cox_strat, exponentiate = TRUE) |> add_global_p(),
    tbl_regression(cox_time_var, exponentiate = TRUE) |> add_global_p()
  ),
  tab_spanner = c("**Standard Cox**", "**Time-varying Cox**")
)
```

Monoclonal gammopathy data Description

Natural history of 1341 sequential patients with monoclonal gammopathy of undetermined significance (MGUS). This is a superset of the mgus data, at a later point in the accrual process

id: subject identifier
age :age at diagnosis, in years
sex: a factor with levels F M
dxyr: year of diagnosis
hgb: hemoglobin
creat:creatinine
mspike:size of the monoclonal serum splike
ptime: time until progression to a plasma cell malignancy (PCM) or last contact, in months
pstat: occurrence of PCM: 0=no, 1=yes
futime:time until death or last contact, in months
death:occurrence of death: 0=no, 1=yes


```{r}
# ─────────────────────────────────────────────────────────────
# Competing Risks Analysis with mgus2
# ─────────────────────────────────────────────────────────────

mgus2
```


```{r}
# --- 1. Prepare mgus2 data ---

# Time-to-first-event:
# If no progression (pstat == 0), use follow-up time (futime)
# If progression occurred (pstat == 1), use progression time (ptime)
mgus2$etime <- ifelse(mgus2$pstat == 0, mgus2$futime, mgus2$ptime)

# Event status:
# 1 = PCM (plasma cell malignancy / progression)
# 2 = Death without PCM (competing event)
# 0 = Censored
mgus2$event <- ifelse(mgus2$pstat == 1, 1,
                      ifelse(mgus2$death == 1, 2, 0))

# Convert to factor — required by tidycmprsk
# Level 0 (first level) = censored reference
mgus2$event_factor <- factor(mgus2$event, levels = 0:2,
                             labels = c("Censored", "PCM", "Death"))

```


```{r}

# ─────────────────────────────────────────────────────────────
# --- 1. Cumulative Incidence Function (CIF) ---
# ─────────────────────────────────────────────────────────────

cuminc_fit <- cuminc(Surv(etime, event_factor) ~ sex, data = mgus2)

cuminc_fit
```


```{r}

# ─────────────────────────────────────────────────────────────
# --- 2. Cumulative Incidence Function (CIF) Plot ---
# ─────────────────────────────────────────────────────────────

ggcuminc(cuminc_fit, outcome = c("PCM", "Death")) +
  add_confidence_interval() +
  add_risktable(risktable_stats = "n.risk") +
  scale_color_manual(values = c("#E63946", "#457B9D")) +
  
  scale_fill_manual(values  = c("#E63946", "#457B9D")) +
  theme_minimal(base_size = 13) +
  labs(
    title    = "Cumulative Incidence of PCM and Death",
    subtitle = "Stratified by sex — mgus2 dataset",
    x        = "Months",
    y        = "Cumulative Incidence",
    color    = "Group",
    fill     = "Group"
  )
```
---

# cohort_data

id: 1 – 200	Unique patient identifier

age: 	Integer	Continuous (years)	Age at cohort entry

sex:	Binary	0 = female, 1 = male

comorbidity:	Binary	0 = absent, 1 = present	Presence of a relevant comorbidity at baseline

trt_start_time:	Integer / NA	Days (1 – 180), or NA	Day on which treatment was initiated; NA if the patient was never treated

fup_time:	Integer	Days (1 – 365)	Total follow-up time — either the day of the event or the day of administrative censoring (Day 365), whichever came first

event:	Binary	0 = censored, 1 = event	Outcome indicator: 1 if the event occurred before Day 365, 0 if administratively censored

```{r}
# ----------------------------------------------------------
# Load data (one row per patient)
# ----------------------------------------------------------

cohort_data <- readRDS("cohort_data.Rds")

# Quick exploration
cohort_data




```
```{r}
table(cohort_data$event)                        # events vs censored

```

```{r}
table(treated = !is.na(cohort_data$trt_start_time))  # treated vs untreated
```


```{r}
# ----------------------------------------------------------
# 1a. Build base_data — adds (tstart, tstop) interval structure
# ----------------------------------------------------------
base_data <- tmerge(
  data1 = cohort_data,
  data2 = cohort_data,
  id    = id,
  event = event(fup_time, event)
)

base_data
```


```{r}
# ----------------------------------------------------------
# 1b. Build td_cox_data — adds time-varying treatment status
# ----------------------------------------------------------
td_cox_data <- tmerge(
  data1      = base_data,
  data2      = cohort_data,
  id         = id,
  trt_status = tdc(trt_start_time)
)

td_cox_data
```


```{r}
# ----------------------------------------------------------
# 2. Extended Cox Model (Time-Dependent Covariate)
# ----------------------------------------------------------
td_cox <- coxph(
  Surv(tstart, tstop, event) ~ trt_status + age + sex + comorbidity,
  data = td_cox_data
)
tbl_regression(td_cox, exponentiate = TRUE) |> 
  add_global_p()
```

---

Data from the European Society for Blood and Marrow Transplantation (EBMT):


id: Patient identification number

prtime: Time in days from transplantation to platelet recovery or last follow-up

prstat: Platelet recovery status; 1 = platelet recovery, 0 = censored

rfstime: Time in days from transplantation to relapse or death or last follow-up (relapse-free survival time)

rfsstat: Relapse-free survival status; 1 = relapsed or dead, 0 = censored

dissub: Disease subclassification; factor with levels "AML", "ALL", "CML"

age: Patient age at transplant; factor with levels "<=20", "20-40", ">40"

drmatch: Donor-recipient gender match; factor with levels "No gender mismatch", "Gender mismatch"

tcd: T-cell depletion; factor with levels "No TCD", "TCD"

```{r}
data(ebmt3)
ebmt3
```
```{r}
tmat <- trans.illdeath(names = c("Transplant", "Recovery", "Relapse/Death"))
tmat
```


prtime: Time in days from transplantation to platelet recovery or last follow-up
prstat: Platelet recovery status; 1 = platelet recovery, 0 = censored

rfstime: Time in days from transplantation to relapse or death or last follow-up (relapse-free survival time)
rfsstat: Relapse-free survival status; 1 = relapsed or dead, 0 = censored

```{r}
# 2. Prepare data in long format
ms_data <- msprep(time = c(NA, "prtime", "rfstime"),
                  status = c(NA, "prstat", "rfsstat"),
                  data = ebmt3, trans = tmat, keep = "age")

ms_data
```

```{r}
# 3. Expand factor covariate — inspect names before using them!
ms_data <- expand.covs(ms_data, "age")

ms_data
```
```{r}
# 4. Fit stratified Cox model
ms_cox <- coxph(Surv(Tstart, Tstop, status) ~
                  age20.40.1 + age.40.1 +
                  age20.40.2 + age.40.2 +
                  age20.40.3 + age.40.3 +
                  strata(trans),
                data = ms_data, method = "breslow")

tbl_regression(ms_cox, exponentiate = TRUE)  |> 
  modify_caption("**Multi-State Transitions: Impact of Age Category (Baseline: <=20)**")
```


---

Rat treatment data from Mantel et al. Three rats were chosen from each of 100 litters, one of which was treated with a drug, and then all followed for tumor incidence.

litter:	litter number from 1 to 100
rx:	treatment,(1=drug, 0=control)
time:	time to tumor or last follow-up
status:	event status, 1=tumor and 0=censored
sex:	male or female

```{r}
data(rats, package = "survival") 
rats
```



```{r}


# 1. Standard Cox (ignores clustering)
cox_naive <- coxph(Surv(time, status) ~ rx + sex, data = rats)

# 2. Mixed-Effects Cox (Gaussian frailty) using coxme package
cox_mixed <- coxme(Surv(time, status) ~ rx + sex + (1 | litter), data = rats)

# Combine for publication table
tbl_merge(
  tbls = list(
    tbl_regression(cox_naive, exponentiate = F) |> 
      add_global_p(),
    tbl_regression(cox_mixed, exponentiate = F) |>
      add_global_p()
  ),
  tab_spanner = c("**Naive (No Frailty)**", "**Mixed Effects (Frailty)**")
)

```

```{r}
# Setup data
data(colon)
colon_death <- subset(colon, etype == 2) # Filter to model death only

# Task 1: Fit the Cox Model
cox_colon <- coxph(Surv(time, status) ~ rx + age, data = colon_death)

# Task 2: Schoenfeld Residuals Test
test_colon <- cox.zph(cox_colon)
print(test_colon) # View p-values

# Task 3: Plot the residuals
ggcoxzph(test_colon, font.main = 12)
# Answer: The p-value for 'rx' < 0.05, and the plot shows a trend 
# line that isn't flat. The PH assumption is violated. The treatment effect 
# is stronger early on and wanes over time.

# Task 4 (Bonus): How to adjust?
# You could use a stratified model if rx was just a confounder, but since it's 
# the treatment of interest, you should use a time-varying coefficient tt(rx) 
# or split the time window (e.g., effect in year 1 vs. effect after year 1).

```

```{r}
# Task 1: Competing Risks with mgus2
# Create binary hemoglobin category
mgus2$hgb_cat <- ifelse(mgus2$hgb > 12, "> 12 g/dL", "<= 12 g/dL")
mgus2$event_factor <- factor(mgus2$ptime, 0:2, c("Censored", "PCM", "Death"))

# Calculate and plot CIF
ex2_cuminc <- cuminc(Surv(futime, event_factor) ~ hgb_cat, data = mgus2)

ggcuminc(ex2_cuminc, outcome = c("PCM", "Death")) +
  add_confidence_interval() +
  add_risktable() +
  theme_minimal() +
  labs(title = "CIF by Hemoglobin Category", x = "Months")

# Task 2: Frailty Modeling with Kidney dataset
data(kidney, package = "survival")

# Fit Naive Cox
naive_kidney <- coxph(Surv(time, status) ~ sex, data = kidney)

# Fit Mixed-Effects (Frailty) Cox using patient ID
mixed_kidney <- coxme(Surv(time, status) ~ sex + (1 | id), data = kidney)

# Compare outputs
tbl_merge(
  tbls = list(
    tbl_regression(naive_kidney, exponentiate = TRUE),
    tbl_regression(mixed_kidney, exponentiate = TRUE)
  ),
  tab_spanner = c("**Naive (Ignores repeated infections)**", "**Frailty (Accounts for Patient ID)**")
)
```

